---
title: "Visualization Fundamentals"
author: "Hyoungwoo Kim"
output: html_document
date: "2025-02-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(nycflights13)
```


```{r}
my_theme <- theme_classic() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 14),
    panel.grid.minor = element_blank()
  )
theme_set(my_theme)
```

## NYC Flights Analysis

### **(a) Create a new column for average flight speed**
```{r}
# Load the NYC flights dataset and select relevant columns
flights_speed <- flights |> 
  select(carrier, air_time, distance) |> 
  mutate(speed = distance / (air_time / 60)) |> # Convert air_time to hours and compute speed
  drop_na() # Remove any rows with missing values

# Display the first few rows
head(flights_speed)
```

### **(b) Analyze variation in flight speed across carriers**
```{r}
# Clean Boxplot without Jitter
speed_plot <- ggplot(flights_speed, aes(x = reorder(carrier, speed, median), y = speed, fill = carrier)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) + 
  coord_cartesian(ylim = c(0, 600)) + # Adjust y-axis range
  labs(
    title = "Variation in Flight Speed Across Carriers",
    x = "Carrier",
    y = "Average Speed (miles per hour)"
  ) +
  theme(legend.position = "none")

# Display the plot
speed_plot

```

## Pokemon Analysis

### **(a) Compute Attack-to-Defense Ratio**
```{r}
pokemon <- read_csv("/Users/hyoungwookim/Desktop/2025 Spring/stat436_s25/poke.csv")

pokemon <- pokemon |> 
  mutate(attack_defense_ratio = Attack / Defense)

head(pokemon)
```

### **(b) Compute Median Attack-to-Defense Ratio per Type**
```{r}
pokemon_median <- pokemon |> 
  group_by(type_1) |> 
  summarise(median_ratio = median(attack_defense_ratio, na.rm = TRUE)) |> 
  arrange(desc(median_ratio))

head(pokemon_median)
```

### **(c) Plot Attack vs. Defense by Type**
```{r}
pokemon_plot <- ggplot(pokemon, aes(x = Attack, y = Defense, color = type_1)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ reorder(type_1, attack_defense_ratio, median)) +
  labs(
    title = "Attack vs. Defense by Pokemon Type",
    x = "Attack",
    y = "Defense"
  ) +
  theme_minimal()

pokemon_plot
```

### **(d) Proposed Interactive Visualization**
- **Question:** How does the attack-defense balance change across different Pokemon generations?
- **Interactive Feature:** Implement a **slider or dropdown menu** to filter Pokemon by generation.
- **Expected Outcome:** Users can explore how Pokemon stats evolved over generations, identifying trends in attack vs. defense.

